Query Optimization for Dynamic Imputation

### 2017-05-15 rebuttal.tex

  \resp{
    We extended our evaluation to include two commonly used imputation methods: mean value imputation and hot-deck imputation.
    These methods are in addition to the original imputation method (CART), which is based on regression trees.
    The regression tree model incorporates correlations between attributes, so should be able to take advantage of any functional dependencies to make better predictions.
    Interpolation based imputation is useful for data where the ordering of tuples is significant, like time series, but is not well suited for survey data where the ordering of tuples is arbitrary.

    We also added details on how to pick the cost functions that define the API for impute operators.
    We provide details on how these functions are defined for all the methods used in our evaluation.

    Our optimizer (like any query optimizer) is sensitive to the quality of its estimates.
    We believe that the cost functions we have chosen lead to good results overall, but there may be cases where our estimates lead to bad outcomes.
    It is an interesting and difficult problem to accurately estimate the quality of a regression model before running the model.
    Any work in this area could be used immediately to improve our algorithm.
  }

  \item W3. Reducing runtime to 10s is excellent, but comes at a very serious accuracy cost (3-5x worse SMAPE). Worse still, there's no way for the user to differentiate between the cases of Q4 and Q8 at runtime. 24\% error relative to direct imputation is huge.

@@

    \resp{We have clarified this in the paper by emphasizing that \textit{Loss} is not a loss in the statistical sense or in the tuple loss sense and by renaming it to $\textsc{Penalty}$.}

  \item S3.4.3: An extremely nice emergent property from the L-value function for the imputation operator seems to be that the operator will be moved to a position in the plan where it has access to the maximum training data. I am slightly concerned that by focusing entirely on size, this might lead to skew in the training set. For example, consider a one-to-many join \verb/µ(R) |X| S/. Your optimization algorithm would likely wish to move imputation up: \verb/µ(R |X| S)/. Yes, there is more training data, but if the fanout for each tuple is skewed, some tuples in R will become over-represented in the join result. A similar observation holds for selection. For example take your Q8, where the conditions "studentdebtowe $>$ 0 AND schooldegree $\geq$ 0" presumably indicate invalid data that should not be used for training. (is this perhaps why you have so much error for full imputation on Q8?)

    \resp{This is a reasonable concern. Though often an appropriate proxy, optimizing for the maximum training set is flawed when the query requires the construction a large, skewed data set. The problem seems to be particular to queries which generate large intermediate tables. In some cases, as when joining two datasets, the intermediate table is a good target for training because it exposes correlations which would be hidden in the separate tables. In other cases, this intermediate table is large and skewed, which causes problems for imputation.

      Avoiding this case requires us to estimate the skew in these intermediate tables, which is an area of active research.
      However, if we had improved intermediate estimates, we could incorporate them to improve our system.
      }
    %\todo{Need a better explanation here.}

@@

  \item D8. Are there impute implementation that are non-blocking? Or do they always have to perform multiple passes over the data?

    \resp{Yes. For example, in time series data, it is reasonable to fill the most recent non-null value forward. This kind of imputation isn't useful for survey data, so we didn't consider it for our experiments. We have extended the experiments section to consider a simple non-blocking imputation method relevant to our dataset: mean value imputation. This is a common technique, which fills in  missing values
    with the average observed value. In our case, we implement a non-blocking variant that estimates the mean from the plan's histograms, avoiding the need to perform a full pass over
    the data before imputing.}

  \item D9. Additional experiments I suggest to include:
    -- Consider a scenario where the user submits multiple similar queries. Show after how many such queries your approach has no longer a competitive advantage compared to performing imputation on the data.
    -- Try different values for $\alpha$ (not just 0 and 1).
    -- Show runtime of enumeration algorithm as queries get more complex.

    \resp{
      It is straightforward to compute the number of queries before we lose our advantage.
      For example, performing full imputation on the ACS dataset takes 355 minutes (Sec. 5.3.2) and we can run a simple query in 4 seconds, so we can run approximately $5,000$ queries before taking more time than full imputation.

      The sensitivity of different values of $\alpha$ depends on the query itself. Performing a scan of values of $\alpha$ with small step sizes can return
      a lot of the same plans for execution. In lieu of this, we extended the results section to include a subsection that presents the final frontiers for
      two queries, one from each dataset used during evaluation. This frontier can be exposed to the user to help in their workflow.

      We explored the planning times for queries with increasing number of joins by running additional experiments with up to 8 joins. 
      The standard algorithm presented in the original version submitted handles
      up to 5 joins, with columns requiring imputation in all tables, with planning time of approximately one second or less. We added a small optimization to help reduce
      planning times for a larger number of joins. We maintain an approximation of the Pareto sets, rather than the exact set, by only adding non-dominated plans that
      are sufficiently different. This optimization reduces planning time linearly with respect to the reduction in the size of the final frontier. Users can increase
      the level of approximation to reduce planning times further.}

### 2017-05-15 paper/algorithm.tex
paper: grammar & spelling

In order to pick a final plan from the frontier, our model introduces a trade-off parameter $\alpha$, which is an upper bound on the $\textsc{Penalty}$ value that the user is willing
to tolerate relative to the minimum possible amongst frontier plans.

@@

The form of $T$ depends on the specific imputation method, but there are several considerations in choosing $T$.
In particular, $T$ should \emph{increase} as the number of tuples and complete and dirty attributes passed to the imputation function increases and it should be take into account the complexity of the imputation algorithm.
Analysis of the time complexity of different imputation algorithms, especially with in-database execution, is beyond the scope of this paper.
As with $P$, it is the ordering between queries generated by $T$ that is more important than the precision of estimates.

@@

\subsubsection{Plan cache}
Before describing the operation of the query planner, we describe the semantics of a specialized data structure for holding sub-plans, called a \emph{plan cache}.
At a high level, a plan cache is a mapping from a set of tables to a set of dirty set-specific dominating plans over these tables.
We store multiple plans rather than a single plan per dirty set, because maintaining a Pareto frontier as a plan progresses through the optimizer's pipeline
allows us to obtain the final set of plans that best trade-off computation
cost and imputation quality. Details of the plan-cache semantics are shown in~\Cref{fig:semantics-plan-cache}.

@@

The plan cache uses the partial order of plans defined by  $\langle \textsc{Dirty}, \textsc{Penalty}, \textsc{Time} \rangle$ to
collect sound and complete Pareto frontiers. Plans with different $\textsc{Dirty}$ sets cannot be compared. The final
Pareto frontier produced by the optimizer corresponds to the collection of plans that best tradeoff imputation quality and
execution time.

>>>>

In order to pick a final plan from the frontier, our model introduces a trade-off parameter $\alpha$, which is an upper bound on the $\textsc{Penalty}$ value that the user is willing
to tolerate relative to the minimum possible among the frontier plans.

@@

The form of $T$ depends on the specific imputation method, but there are several considerations in choosing $T$.
In particular, $T$ should \emph{increase} as the number of tuples and complete and dirty attributes passed to the imputation function increases and it should take into account the complexity of the imputation algorithm.
Analysis of the time complexity of different imputation algorithms, especially with in-database execution, is beyond the scope of this paper.
As with $P$, it is the ordering between queries generated by $T$ that is more important than the precision of estimates.

@@

\subsubsection{Plan cache}
Before describing the operation of the query planner, we describe the semantics of a specialized data structure for holding sub-plans, called a \emph{plan cache}.
At a high level, a plan cache is a mapping from a set of tables to a set of dirty set-specific dominating plans over these tables.
We store multiple plans rather than a single plan per dirty set, because maintaining a Pareto frontier as a plan progresses through the optimizer's pipeline
allows us to obtain the final set of plans that best trade-off computation
cost and imputation quality. The plan-cache semantics are shown in~\Cref{fig:semantics-plan-cache}.

@@

The plan cache uses the partial order of plans defined by  $\langle \textsc{Dirty}, \textsc{Penalty}, \textsc{Time} \rangle$ to
collect sound and complete Pareto frontiers. Plans with different $\textsc{Dirty}$ sets cannot be compared. The final
Pareto frontier produced by the optimizer is the set of plans with the best imputation quality-execution time trade-offs.

### 2017-05-15 paper/examples.tex
paper: grammar & spelling

>>>>

\review{
As an alternative strategy, the analyst can start by executing their query with $\alpha=1$ and incrementally reducing $\alpha$, as long as execution of the query completes within some user-defined time bound.
\ProjectName{}'s planning algorithm can expose the final set of plans with optimal time-quality trade-offs, which allows the user to adjust their choice of $\alpha$ to select from these plans.
}

### 2017-05-15 paper/experiments.tex
old ref
paper: grammar & spelling

\todo{Distinguish between underlying imputation methods}
For $\alpha=0$ and $\alpha=1$, we pair up tuples and compute SMAPE for all queries. That is,
we compute tuple-wise absolute percentage deviations within each iteration of a query, and
we report this value averaged over all iterations.  We
can see that optimizing for quality indeed reduces the SMAPE of query results.  In general, the SMAPE relative 
to the baseline is low---\todo{update: between \lowsmapealphazero{} and
\highsmapealphaoneexacs{} percent}---indicating that on-the-fly imputation produces results similar to the baseline. Notably, the results with largest error correspond
most closely to one existing approach of dropping all \nullv{}s.

@@

\review{
\Cref{fig:pareto-frontiers} shows the final Pareto frontiers produced by the planner for \Cref{q1} and \Cref{q6}. These are the resulting set of plans after pruning dominated plans throughout our algorithm.
and encapsulate the possible tradeoffs that the user can make in choosing a final plan for execution. The frontier is available at the end of the planning stage and
can be exposed to the user to help in guiding their workflow.
}

@@

\begin{figure}
\centering
\includegraphics[scale=0.4]{figures/pareto_frontiers_plot.png}
\caption{\review{
    \ProjectName{}'s final planner Pareto frontier for \Cref{q1} and \Cref{q6}. 
    For clarity, we have removed from the plot plans that provide nearly indistinguishable tradeoffs in either dimension.
}}
\label{fig:pareto-frontiers}
\end{figure}

@@

\review{
However, \ProjectName{}'s planning algorithm is still exponential in the number of joins in the plan, and the use of Pareto sets exacerbates this exponential growth.
We explored the limits of our planning algorithm by constructing a series of queries with increasing number of joins over the CDC tables.
Planning times (~\Cref{fig:planning-times}) for queries involving 1--5 joins---a practical value for  real world exploratory queries---are less than 1 second.
We extended \ProjectName{} to support an approximation of Pareto sets to improve planning times for queries with 6 or more joins. Approximate Pareto sets
displayed a reduction in planning time linear with respect to the reduction in size of the final frontier. For queries with 6--8 joins, approximate sets achieved a reduction
in planning time between 14\% and 23\%, on average. Users can increase the level of approximation to reduce planning times further.
}
%\todobox{Jose: I'm not a fan of this line, I would remove}{Furthermore, at that point, it may make more sense to specify a coherent model over the entire base table instead.}

>>>>

\todo{Distinguish between underlying imputation methods}
For $\alpha=0$ and $\alpha=1$, we pair up tuples and compute SMAPE for all queries. That is,
we compute tuple-wise absolute percentage deviations within each iteration of a query, and
we report this value averaged over all iterations.  We
can see that optimizing for quality indeed reduces the SMAPE of query results.  In general, the SMAPE relative 
to the baseline is low---\todo{update: between \lowsmapealphazero{} and
\highsmapealphaoneexacs{} percent}---indicating that on-the-fly imputation produces results similar to the baseline. Notably, the results with the largest error correspond
most closely to one existing approach of dropping all \nullv{}s.

@@

\review{
\Cref{fig:pareto-frontiers} shows the final Pareto frontiers produced by the planner for \Cref{q1} and \Cref{q6}. These are the resulting set of plans after pruning dominated plans throughout our algorithm.
and encapsulate the possible trade-offs that the user can make in choosing a final plan for execution. The frontier is available at the end of the planning stage and
can be exposed to the user to help in guiding their workflow.
}

@@

\begin{figure}
\centering
\includegraphics[scale=0.4]{figures/pareto_frontiers_plot.png}
\caption{\review{
    \ProjectName{}'s final planner Pareto frontier for \Cref{q1} and \Cref{q6}. 
    For clarity, we have removed from the plot plans that provide nearly indistinguishable trade-offs in either dimension.
}}
\label{fig:pareto-frontiers}
\end{figure}

@@

\review{
However, \ProjectName{}'s planning algorithm is still exponential in the number of joins in the plan, and the use of Pareto sets exacerbates this exponential growth.
We explored the limits of our planning algorithm by constructing a series of queries with increasing number of joins over the CDC tables.
Planning times for queries involving 1--5 joins---a practical value for  real world exploratory queries---are less than 1 second.
We extended \ProjectName{} to support an approximation of Pareto sets to improve planning times for queries with 6 or more joins. Approximate Pareto sets
displayed a reduction in planning time linear with respect to the reduction in size of the final frontier. For queries with 6--8 joins, approximate sets achieved a reduction
in planning time between 14\% and 23\%, on average. Users can increase the level of approximation to reduce planning times further.
}
%\todobox{Jose: I'm not a fan of this line, I would remove}{Furthermore, at that point, it may make more sense to specify a coherent model over the entire base table instead.}

### 2017-05-15 paper/relatedwork.tex
paper: grammar & spelling

\review{
\textcite*{yang2015lenses} develops an incremental approach to ETL data cleaning, through the use
of probabilistic compositional repair operators, called Lenses. These can be used
to repair missing values by imposing data constraints such as \texttt{NOT NULL}. 
Similarly to \ProjectName{}, they provide a way to balance between the quality of the cleaning 
performed (expressed by the confidence of their output) and the cost of those operations. However, their 
repairs take place on a per-tuple basis, with the non-deterministic layer of their system replacing place-holder
variables with inferred values. In contrast, \ProjectName{} optimizes imputation at the plan-level and
provides the same impute technique for all tuples streaming through a given operator.
}

@@

\subsection{Forecasting and Databases}
\textcite*{parisi2011embedding} introduces the idea of incorporating time-series forecast operators into
databases, along with the necessary relational algebra extensions. Their work explores the theoretical
properties of forecast operators and generalizes them into a family of operators, distinguished by
the type of predictions returned. They highlight the use of forecasting for replacing missing values.
In subsequent work~\cite{parisi2013temporal}, they identify various equivalence and containment
relationships when using forecast operators, which could be used to perform query plan transformations that guarantee the same result. They
explore forecast-first and forecast-last plans, which perform forecasting operations before and after executing all traditional
relational operators, respectively.

@@

Similarly to Fa, \ProjectName{} provides a declarative language, as
a subset of standard SQL.\@ \ProjectName{}, however, is not searching the space of possible
imputation models, but rather the space of query plans that incorporate imputation operators. Another major point
of distinction between \ProjectName{} and Fa is that the latter doesn't consider trade-offs between accuracy and computation time, but rather returns the most accurate forecast (with some stopping criterion).

>>>>

\review{
\textcite*{yang2015lenses} develops an incremental approach to ETL data cleaning, through the use
of probabilistic compositional repair operators, called Lenses. These can be used
to repair missing values by imposing data constraints such as \texttt{NOT NULL}. 
As with \ProjectName{}, they provide a way to balance between the quality of the cleaning 
performed (expressed by the confidence of their output) and the cost of those operations. However, their 
repairs take place on a per-tuple basis, with the non-deterministic layer of their system replacing place-holder
variables with inferred values. In contrast, \ProjectName{} optimizes imputation at the plan-level and
provides the same impute technique for all tuples streaming through a given operator.
}

@@

\subsection{Forecasting and Databases}
\textcite*{parisi2011embedding} introduce the idea of incorporating time-series forecast operators into
databases, along with the necessary relational algebra extensions. Their work explores the theoretical
properties of forecast operators and generalizes them into a family of operators, distinguished by
the type of predictions returned. They highlight the use of forecasting for replacing missing values.
In subsequent work~\cite{parisi2013temporal}, they identify various equivalence and containment
relationships when using forecast operators, which could be used to perform query plan transformations that guarantee the same result. They
explore forecast-first and forecast-last plans, which perform forecasting operations before and after executing all traditional
relational operators, respectively.

@@

As with Fa, \ProjectName{} provides a declarative language, as
a subset of standard SQL.\@ \ProjectName{}, however, is not searching the space of possible
imputation models, but rather the space of query plans that incorporate imputation operators. Another major point
of distinction between \ProjectName{} and Fa is that the latter doesn't consider trade-offs between accuracy and computation time, but rather returns the most accurate forecast (with some stopping criterion).
